---
title: ggplot and other tools
subtitle: "Bio302 - Biological data analysis II"
author: 
- name: "Richard J. Telford"
institute: Department of Biology, University of Bergen
date: '2016-05-05'
output: 
 ioslides_presentation:
  widescreen: true
---

```{r setup_gg, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 5, cache = TRUE, message = FALSE, cache.lazy = TRUE)
library("dplyr")
```

## ggplot2

Part of `tidyverse`

```{r}
library("tidyverse")
```
Produces publication quality graphics

Implementation of the Grammar of Graphics

Written by Hadley Wickham


## Some Data

Population, life expectancy and per capita GDP by country since the 1950s

```{r}
if(!require("gapminder")){
  install.packages("gapminder")
  library("gapminder")
}
data("gapminder")
head(gapminder)
```

## Anatomy of a ggplot2 command

Add components together.
```{r anatomy}
g <- ggplot(gapminder, #data
       aes(x = gdpPercap, y = lifeExp, size = pop, colour = continent)) + #aesthetics
  geom_point() + # geometry
  #geom_smooth() + #statistics
  facet_wrap(~ year) + #facets
  #coord_flip + #coordinates
  scale_x_log10() + #scales
  labs(x = "GDP per capita") + #labels
  theme() #theme
```

Can add components to existing ggplot2 object
```{r, eval = FALSE}
g <- ggplot(gapminder, #Data
       aes(x = gdpPercap, y = lifeExp, size = pop, colour = continent)) #aesthetics
g <- g +  geom_point() 
```

##
Need to `print` ggplot2 object
```{r, fig.height=5, fig.width=5.5}
print(g)
```


## Data
Data should be a tidy `data.frame` or `tibble`

Variables in columns, and observations the rows

May need to reshape data with `tidyr::pivot_longer`

May need to subset or summarise data

## Aesthetics

How to map the columns in the data onto the visual properties of the plot

`aes(x = gdpPercap, y = lifeExp, size = pop, colour = continent)`

Other aesthetics include 

 - `linetype`
 - `alpha`
 - `fill` # for areas
 - `ymax`
 - `ymin`
 - `group` # to denote how data should be grouped


## Geometry

What to draw. Can use more than one.

Use different aesthetics

Some examples

 - `geom_bar()`
 - `geom_histogram()`
 - `geom_line()`
 - `geom_path()`
 - `geom_histogram()`
 - `geom_text()`
 - `geom_tile()`

Every geom has an associated stat. Can override default.

##
```{r}
g <- ggplot(filter(gapminder, year == 1957), aes(x = lifeExp, fill = continent)) +
  geom_histogram()
g
```

## Setting vs mapping aesthetics

Map aesthetics inside `aes`
```{r, fig.height=3, fig.width=3}
ggplot(filter(gapminder, year == 1957), 
       aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point()
```

##

Try to make all points blue

```{r, fig.height=3, fig.width=3}
ggplot(filter(gapminder, year == 1957), 
       aes(x = gdpPercap, y = lifeExp, colour = "blue")) +
  geom_point()
```

##

Set aesthetics outside `aes` 
```{r, fig.height=3, fig.width=3}
g2 <- ggplot(filter(gapminder, year == 1957), aes(x = gdpPercap, y = lifeExp)) +
  geom_point(colour = "blue")
g2
```


## Statistics

Statistics for the plot

- `stat_bin()`
- `stat_histogram()`
- `stat_boxplot()`

Every stat has an associated geom. Can override default.

## geom_smooth()

Adds regression model + 95% CI to plot

For small data sets, uses loess

For large data sets, uses gam

Can override and use any `method`

Works best for simple models. Fit complex models seperately

Uses `x`, `y`, weight from `aes`

`g + geom_smooth(method = "lm", formula = y ~ x)`


## Scales

Control how the data are mapped onto the aesthetics

 - `scale_x_continuous`
 - `scale_fill_discrete`
 - `scale_colour_brewer`
 - `scale_y_log10`

Most useful arguments: `breaks`, `labels` and `values`

##
```{r}
g <- g + scale_fill_manual(values = continent_colors) 
g
```

##

```{r}
g2 + scale_x_log10()
```


## Facets

Split the data into partitions

 - `facet_wrap()` By one variable
 - `facet_grid()` By two variables

```{r, fig.height = 3.5}
g2 + facet_wrap(~ continent)

```


## Coordinates

 - `coord_flip()` Flip x and y axis
 - `coord_map()` Projections for maps from `mapproj`
```{r}
g + coord_flip()
```


## Labels etc

If the names of your data are not the labels you need
```{r}
g <- g + labs(x = "Life Expectancy", y = "Count", fill = "Continent")
g
```



## Themes

Don't like the default plot? Change it with theme

 - `theme_bw()`
 - `theme_classic()`
```{r, fig.height=3, fig.width=4}
g + theme_classic()
```

 
##

Change any element of the figure

```{r, fig.height = 3.5}
g <- ggplot(gapminder, aes(x = continent, y = lifeExp)) + 
  geom_boxplot()
g + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

See `?theme` to see elements to change
 
## Legends

Legends appear automatically when required

Can turn off with `show.legend` argument to the `geom_xxx`

Can move with

`g + theme(
     legend.justification = c(1, 0), 
     legend.position = c(1, 0))`
 
## Saving plots
```{r}
ggsave("myplot.png")
```
Works out what format to use from extension

Arguments to change size


## Multiple plots per page
Need help from extra packages `grid`, `gridExtra`, or `cowplot`

```{r}
library(gridExtra)
grid.arrange(g, g2, ncol = 2, nrow = 1)
```

## Resources

GGplot cheat sheet https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

R Graphics Cookbook - http://www.cookbook-r.com/Graphs/index.html
  
ggplot2: Elegant Graphics for Data Analysis - https://ggplot2-book.org/

Stackoverflow.com search with tag [ggplot2]

## GG related packages

ggfortify - plotting R objects from some packages

GGalley - pairs plots

ggmaps - maps

plotly - interative plots

cowplot - different default themes

## GGally{.columns-2}

Pairs plots

```{r, fig.height = 5, fig.width = 5}
if(!require("GGally")){
  install.packages("GGally")
  library("GGally")
}

ggpairs(
  gapminder,
  columns = 2:ncol(gapminder),
  mapping = aes(colour = continent)
  )
```

## ggmaps

Use Google Maps as background

Need a free API key 

```{r ggmap, message = FALSE, eval = TRUE}
library(ggmap)
dat <- tibble(Long = 18.5, Lat = -34)
map <- get_googlemap(center = c(lon = 18.5, lat = -34), zoom = 11)
g <- ggmap(map) +
  geom_point(aes(x = Long, y = Lat), data = dat, size = 2) +
  labs(x = "Longitude °E", y = "Latitude °N")
```

`center` can be vector giving longitude and latitude or an address

##
```{r print-ggmap, echo = FALSE, eval = TRUE}
print(g)
```

Can add points, lines, polygons, etc.

__Make a satellite map of your field area region__

##Don't like cluttered google maps?

Turn off selected elements
```{r clean-ggmap, eval = FALSE, tidy=TRUE}
map <- get_googlemap(
  center = c(lon = 18.5, lat = -34), 
  zoom = 10, 
  maptype = "terrain", 
  language = "en-EN", 
  style = 'feature:road|visibility:off&
    style=feature:administrative.locality|element:labels|visibility:off&
    style=feature:landscape|visibility:on|saturation:-100|lightness:50&
    style=feature:landscape|element:labels|visibility:off&
    style=feature:poi|visibility:off&
    style=feature:transit|visibility:off'
)

ggmap(map)
```

See https://developers.google.com/maps/documentation/static-maps/
