---
title: "Palaeoenvironmental reconstructions"
author: "Richard J. Telford"
date: "October 14, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE)
```

## Setup

Transfer functions for palaeoenvironmental reconstructions can be fitted with the `rioja` or `analogue` packages in R. Other packages such as `vegan` and `palaeoSig` can help interpretation.

Download and install these packages with `install.packages()`

```{r, eval = FALSE}
# install.packages("rioja")
# install.packages("analogue")
# install.packages("palaeoSig")
# install.packages("vegan")
# install.packages("tidyverse")#for ggplot etc
# install.packages("devtools")
# devtools::install_github("gavinsimpson/ggvegan")
# devtools::install_github("richardjtelford/ggpalaeo")
```
This needs doing only once. The packages need loading into R every session with the command `library`.

```{r}
library("rioja")
library("palaeoSig")
library("tidyverse")
library("ggvegan")
library("ggpalaeo")
```

Packages `rioja` and `analogue` have functions with the same names - don't load both at the same time.

```{r}
set.seed(42)#ensures results are same every time
```
## Data

For convenience, we are going to use some of the datasets available within the `rioja` package.

```{r}
#NB any text after a '#' is a comment and ignored by R 
#get help on any function with '?function_name' 
data("SWAP")#SWAP calibration set
data("RLGH")#fossil data from the Round Loch of Glenhead
```

This has imported the data in lists. We can break these up into their component parts to simplify the code later.

```{r}
# '<-' assigns the result on the right to the name on the left
# '$' extracts a named component of a list
spp <- SWAP$spec
pH <- SWAP$pH
spNames <- SWAP$names

fos <- RLGH$spec
ageDepth <- RLGH$depths
```

## Species-environement relationships 

Plot calibration set species abundances against pH. Add a smoother to highlight the main trends.

```{r}
ggplot(spp/100, aes(x = pH, y = TA003A)) + 
  geom_point() +
  geom_smooth(method = "gam", method.args = list(family = "quasibinomial"))

```

The relationship between all species and the environment can be seen with an ordination.

```{r}
ord <- cca(sqrt(spp) ~ pH)
ord
autoplot(ord)
#learn more about ordinations in Bio303
```


## Fit a transfer function


```{r, eval = FALSE}
mod <- WA(y = sqrt(spp), x = pH)
#crossvalidate
mod <- crossval(mod)
performance(mod)
```
```{r, echo = FALSE, include = FALSE}
mod <- WA(y = sqrt(spp), x = pH)
#crossvalidate
mod <- crossval(mod)
```
```{r, echo = FALSE}
performance(mod)
```
**Question:** do any of the other transfer function methods perform better - try WAPLS, MAT, MLRC? Look only at the cross-validated performance. 

By default `crossval` uses leave-one-out cross-validation. Bootstrap cross-validation is also useful as it gives sample specific errors. Other forms of cross-validation are appropriate if there is pseudoreplication or autocorrelation.

## Plot a stratigraphic plot of the fossil data

```{r}
strat.plot(d = fos, yvar = ageDepth$Age, scale.percent = TRUE, y.rev = TRUE)
```

## Apply the transfer function

```{r}
recon <- predict(mod, sqrt(fos))
tibble(age = ageDepth$Age, pH = recon$fit[, "WA.inv"]) %>% 
  ggplot(aes(x = age, y = pH)) + 
  geom_line()
```

## Diagnostics & checks

Several diagnostics available - goodness of fit, distance to closest analogue - in `analogue` package.

Timetrack of fossil samples in calibration set ordination
```{r}
#use 'analogue::' to access functions without loading library - prevents name conflicts with rioja
tt <- analogue::timetrack(X = sqrt(spp), passive = sqrt(fos), env = pH) 
plot(tt)
```

Residual length to constraining ordination axis
```{r}
rlen <- analogue::residLen(X = sqrt(spp), env = pH, passive = sqrt(fos))

plot(rlen)

autoplot(rlen, df = ageDepth, x_axis = "Age") +
labs(x = "Age", y = "Squared residual distance", fill = "Goodness of fit")
```
Colours show how good the fit is.

Analogue distances

```{r}
AD <- analogue_distances(spp, fos)
autoplot(AD, df = ageDepth, x_axis = "Age") +
 labs(y = "Squared-chord distance", x = "Age")
```


Significance tests in `palaeoSig`.

```{r}
tst1 <- randomTF(spp = sqrt(spp), env = pH, fos = sqrt(fos), n = 99, col = 1, fun = WA)
tst1$sig
autoplot(tst1)
```

## Further reading

Juggins and Birks (2012) Quantitative Environmental Reconstructions from Biological Data. In _Tracking Environmental Change Using Lake Sediments, Data Handling and Numerical Techniques_ https://www.springer.com/gp/book/9789400727441



