---
title: "Regular expressions for strings"
author: "Richard J. Telford"
date: "September 29, 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Strings

We might want to 

- search a character vector for strings with a certain pattern
- replace a pattern
- extract a pattern

Regular expressions can help with these

Use base R or `stringr` package

## Detecting a pattern

Very useful inside `dplyr::filter`
```{r}
a <- c("palaeoecology", "paleoecology", "Palaeoenvironment")
grepl(pattern = "palaeo", x = a, ignore.case = TRUE)

#or use stringr package
stringr::str_detect(string = a, pattern = "palaeo")#no case option
```

## Replacing a pattern

Very useful with `dplyr::mutate` for cleaning data
```{r}
a <- c("palaeoecology", "paleoecology", "Palaeoenvironment")
gsub(pattern = "paleo", replacement = "palaeo", x = a, ignore.case = TRUE)

#or use stringr package
stringr::str_replace_all(string = a, 
                         pattern = "paleo", 
                         replacement = "palaeo")
```

## Extracting a pattern

```{r}
a <- c("palaeoecology", "paleoecology", "Palaeoenvironment")


# stringr package
stringr::str_extract(string = a, pattern = "[pP]ala?eo")
```

## Regular expressions{.smaller}

- `.` any character
- `\\.` an actual dot. The `\\` escape the `.`
- `[a-zA-Z]` any letter upper or lower case
- `\\d` a number

### Anchors

- `^`The start of the string
- `$` The end of the string

### Number
- `a?` maybe one "a"
- `a+` at least one "a"
- `a*` zero or more "a"
- `a{3, 5}` three - five "a"


## Example

Exacting the core name from the file name
```{r}
path <- "data/GeoB8323-2_pollen_analysis.tab"
stringr::str_extract(path, "GeoB\\d{4}-\\d")
```
## RVerbalExpressions

Makes writing complex regular expressions easy.

New package  - not yet on CRAN
```{r}
devtools::install_github("VerbalExpressions/RVerbalExpressions")
```

```{r}
library("RVerbalExpressions")
expr <- rx() %>% 
  rx_find("GeoB") %>% 
  rx_digit() %>% 
  rx_multiple(min = 4, max = 4) %>% 
  rx_find("-") %>% 
  rx_digit()
expr

stringr::str_extract(path, expr)

```

## Task

The column names on `GeoB8323_1` are awkward

Remove " [#/cm**3]" and " [#]" from the column names with one regular expression

Use `names()` (or tidyverse solution with `rename_all`)

```{r}
names(GeoB8323_1) <- gsub(..., x = names(GeoB8323_1))
```
replace the ... with your code

