---
title: "Rmarkdown"
author: "Richard J. Telford"
date: "January 30, 2019"
output: ioslides_presentation
bibliography: etc/reproducibility.bib
csl: etc/elsevier-harvard_rjt.csl
---

```{r setup_rmd, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
bt <- "`"
```

## Why

 - Interlaced R code and text compiles to make presentation/paper/thesis 
 - Dynamic documents - update when data/code changes
 - Reproducible analyses
 - Simple formatting
 - Text files - good for version control

## Output formats

Documents

 - PDF
 - HTML
 - word

Presentations

 - ioslides
 
 PDF needs extra packages and TeX `tinytex::install_tinytex()`

## Getting stated

In RStudio

`File>>New File>>R Markdown...`

Select options for document format

R gives example document to edit

Click `Knit` button to render

Or use `rmarkdown::render()`

## YAML

Meta data at top of Rmd file that describes how file is to be processed.

```
---
title: "Untitled"
author: "Richard J. Telford"
date: "`r paste0(bt, "r format(Sys.Date(), '%d %B %Y')", bt)`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: pdf_document
classoption: a4paper
---
```


## R chunks

R code in chunks

\```{r chunkName, eval = TRUE}  
#some R code  
library(vegan)  
x <- 2  
y <- 7  
\```

Chunk name is optional.

Chunk options

 - `eval` Run the R code
 - `echo` Show the R code

Full list [http://yihui.name/knitr/options/](http://yihui.name/knitr/options/)

## Inline code

Seven times six is `` `r '\x60r 7 * 6\x60'` ``

Seven times six is `r 7 * 6`



## Formatting

```
# title (new section in ioslides)
## subtitle (new page in ioslides)
### sub-subtitle
#### sub-sub-subtitle
_italics_ or *italics*
__bold__ or **bold**
 $A = \pi*r^{2}$ 
```
<hr>
### sub-subtitle  
_italics_ or *italics*  
__bold__ or **bold**  
 $A = \pi r^{2}$  
 <hr>

[Rmarkdown cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)

Can include LaTeX commands (for pdf) or html (for html)

## Tables {.subsection}

```{r tables, echo = TRUE}
head(iris, 3) %>% knitr::kable(caption = "A kable table")
```

`kableExtra` package for more control

## Table and Figure numbering {.subsection}

```{r comment = "", echo = FALSE}
'{r cars-figure, fig.cap="Plot showing relationship ..."}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
' -> my_code_string

cat("```", my_code_string, "```", sep = "")
```

Refer to figure with \\\@ref(fig:cars-figure)) 

Important: 

- no `_` in chunk name
- use `bookdown::pdf_book` as output format in YAML

Possible to reference external files with LaTeX `xr` package.

## Citations {.subsection}

In the YAML

```
bibliography: reproducibility.bib
csl: elsevier-harvard.csl
```
- bibliography is bibtex file
- csl file sets citation style

Many csl files [available]( 
https://github.com/citation-style-language/styles)


## Citations

In the text

```
Reproducibility code [@Cooper2017] ...
Cooper and Hsing [-@Cooper2017] discuss ... 
```

Output depends on csl file

<hr>
Reproducibility code [@Cooper2017] ...   
Cooper and Hsing [-@Cooper2017] discuss ...  
<hr>

Reference list printed at end of the MS.

`citr` package makes Rstudio add-in for adding citations to the text.

## Journal specific formatting {.subsection}

Use `rticles` package to produce MS in journal specific format. 

To make draft with correct YAML. Many different journal styles available

```{r rticles, eval = FALSE, echo = TRUE}
library(rmarkdown)
draft("MyArticle.Rmd", template = "plos_article", package = "rticles")
```

## Code overload?

Slow code? Cache chunks with 

```{r comment = "", echo = FALSE}
'{r chunk, cache = TRUE}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
' -> my_code_string

cat("```", my_code_string, "```", sep = "")
```

Too much code? Use `knitr::read_chunk` to read code from a file.

```{r comment = "", echo = FALSE}
'{r load_chunks, cache = FALSE, message=FALSE}
knitr::read_chunk("R/load_data.R")
# file includes line  ## ---- load_data
' -> code

'{r load_data}' -> code2

cat("```", code, "```\n```", code2, "\n```", sep = "")

```

Better solution - `drake` package

## drake

Caches intermediate objects, only remakes them if code or data changes.

See example

```{r, eval = FALSE}
drake::drake_example("main")
```



## Reading list

Reproducible Science with Rmarkdown
http://pakillo.github.io/Rmarkdown_talk_SevillaR_Nov2014/#/

Rmarkdown reference guide https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf

Rmarkdown cheat sheet
https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf
